---
title: "Visualization in R: Mistakes, Fixes and Tools"
author: "Nick Strayer"
date: "May 17, 2016"
output:
 html_document:
   theme: paper
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
knitr::opts_chunk$set(fig.align = "center")
knitr::opts_chunk$set(message=FALSE)
```

---

# Outline

- [What to use?](#what-to-use)
- [Common Mistakes/fixes](#common-mistakes)
    - [Pie Charts](#pie-charts)
    - [Bar Charts](#bar-charts)
    - [Box Plots](#box-plots)
    - [Word Clouds](#word-clouds)
    - [Truncated Axes](#truncated-axes)
    - [Multiple Axes](#multiple-axes)
    - [Information Overload](#information-overload)
    - [The Third Dimension](#third-dimension)
- [Ggplot](#ggplot)
    - [Strengths](#strengths)
    - [Weaknesses](#weaknesses)
    - [Bottom Line](#summary)
- [New Tools/Libraries](#other_libraries)
    - [plotly](#plotly)
    - [html widgets](#html_widgets)
      - [Manhattan Plots](#manhattan_plot)
- [Complex Visualizations/Dashboards](#complex_viz)
    - [Ggpairs](#ggpairs)
    - [Plan Ahead](#requirements_analysis)
    - [Don't go crazy](#advice)
- [Beauty Vs. Clarity](#beauty_clarity)

---


# <a name="what-to-use"></a> What to use?

> "The best camera is the one that's with you." –Chase Jarvis.

A very common question for people starting out with R and visualization is "which library should I use?" Like most things there is no right answer. Every situation is different. I break it down to the following four points.

- Jeff Leek has a [fantastic article](http://simplystatistics.org/2016/02/11/why-i-dont-use-ggplot2/) on his blog about this issue.

- Ultimately it comes down to what you know. You can do an absolutely amazing amount in most tools (even excel) so do what you like best.

- For most people in R the choice is Ggplot vs Base. I mostly use Ggplot because it's what I am the most familiar with (and it has nice defaults (more on this later)).

- Whatever you choose will in the not to distant future be old and replaced by the new best thing, so understanding the concepts is a much better investment of time.

---

# <a name="common-mistakes"></a>Common Mistakes

A lot of data visualization is common sense, but some of it isn't. These are a few of the examples of mistakes made very frequently by visualization practitioners.

## <a name="pie-charts"></a> The Pie Chart

Okay, let's get the elephant out of the room first. The pie chart elicits a similar response in a data-viz person as a computer scientist's prediction algorithm to a statistician. Initially claims of blaspheme but sometimes upon closer inspection grudging respect.

```{r}
# a simple pie chart
data = data.frame(
  val  = c( 8 ,  6 ,  9 ,  4 ,  2 , 3.5),
  labs = c("a", "b", "c", "d", "e", "f") )

pie(data$val, data$labs)
```

So why all the ire?

Humans have a very hard time interpreting angles, and that's how a pie chart encodes the data. Looking at the code/chart above we know that d and f are 0.5 apart, or f is only 87.5% of the size of d, but upon initial inspection the average user would probably say they are the same.

So let's fix it.
```{r}
library(ggplot2)
ggplot(data, aes(y = val, x = labs)) + geom_bar(stat = "identity")
```

Using a bar chart we can clearly see that f is smaller than d.

We can also use something called a tree map: 

```{r}
library(treemap)

treemap(data, c("labs"), "val")
```

I would argue this is actually worse than the pie chart, but if it is certianly a good option for some types of data. Even simpler you can simply do a stacked bar chart. 

```{r}
ggplot(data, aes(1, val, fill=labs, width=0.2)) + geom_bar(stat="identity")
```

Out of all of these options a bar chart is very arguably the most clear. 

Is all the hate warranted for pie charts?

Penn postdoc Randal Olsen has a good [blog post on pie charts.](http://www.randalolson.com/2016/03/24/the-correct-way-to-use-pie-charts/). It is a highly recommended read but to directly quote his rules on pie charts:

1. The parts must sum to a meaningful whole. Always ask yourself what the parts add up to, and if that makes sense for what you’re trying to convey.
2. Collapse your categories down to three or fewer. Pie charts cannot communicate multiple proportions, so stick to their strengths and keep your pie charts simple.
3. Always start your pie charts at the top. We naturally start reading pie charts at the top (the 0° mark). Don’t violate your reader’s expectations.

People intuitively get pie charts so don't rule out their use entirely, but make sure you are using them properly.

---



##<a name="bar-charts"></a> When To Use a Bar Chart

Bar charts are fantastic tools. It seems that more often than not they are the best visualization for the job, often out-competing more complicated flashy visualizations in terms of ease of reading/comprehension. There are some instances where they are not appropriate however.

As a general rule of thumb if the measure is a quantity of something then it makes sense to use a bar chart. This would include number of infections, a person's weight etc..

In this example we will look at a group of patients and their percentiles for vitamin d levels in their blood. 

First we plot with a bar plot.

```{r}
data = data.frame(student = c("Tina", "Trish", "Kevin", "Rebecca", "Sarah"),
                  percentile = c(25, 95, 54, 70, 99)  ) #percentile of d levels

plot = ggplot(data, aes(x = student, y = percentile))

plot + geom_bar(stat = "identity")
```

So the hierarchy of the data is clearly visible but the intuitive interpretation of the bar is slightly confusing. A percentile is not a sum of values but simply a place on the continuum of a scale. In addition we have a tendency to assign good or bad to large or small levels of bar charts when in this case the middle would be best.

Now we visualize it as a dot-plot.

```{r}
# Hacking the geom_pointrange a bit so that the lines are the whole width.
plot + geom_pointrange(aes(ymin = 0, ymax = 100)) + coord_flip()
```

This is more legible and intuitive. We see that the measure is simply a point where the student falls, not the accumulation of percentiles.

There are some exceptions to this rule. For instance: weight being looked at for a single person over time might be best shown on a line chart. Like almost everything in visualization thinking carefully about what your data are before plotting them is important.



---

## <a name="box-plots"></a> Box Plots

Box plots are, like the pie chart, one of the first visualization techniques we are taught. However, it is not necessarily a good one and many better new options have arisen.

The problem with box plots is they obscure trends more subtle than the quantiles. Take for instance the following two box plots: 

```{r, echo=FALSE}
#boxplot code

#generate string of values. 
vals <- runif(100, min = 10, max = 50)
vals <- sort(vals)

min <- 10 #   1
q1 <-  15 #   25
med <- 23 #   50
q3 <-  34 #   75
max <- 42 #   100

vals1 = 1:101
#quantile(vals1, c(0,0.25, 0.5, 0.75, 1))

vals2 = vals1
vals2[2:25] = 15
vals2[27:50] = 42
vals2[52:75] = 63
vals2[77:100] = 82
#quantile(vals2, c(0,0.25, 0.5, 0.75, 1))

data = rbind(data.frame(val = vals1, dataset = "one"),
             data.frame(val = vals2, dataset = "two"))
```


```{r}
p <- ggplot(data, aes(dataset, val))
p + geom_boxplot(fill = "steelblue", color = "grey") + labs(title = "Box Plots")
```

Given the information that we are provided with a standard box plot we would say that these groups are identical. They have the same quantiles that the box plot looks at. 

What happens if we try another way of visualizing the distribution of data? 

First we look at the dot plot:

```{r}
p + geom_dotplot(binaxis = "y", stackdir = "center", 
                 fill = "steelblue", color = "steelblue") + 
  labs(title = "Dot Plots")
```

Now we can see that these data are very differently distributed. 

Another method of visualizing the distribution of the groups is a violin plot. This is essentially a kernel density plot of the previous plot. Useful for when the data are very large and a dot plot is not particularly useful due to the large number of dots drawn. 

```{r}
p + geom_violin(adjust = .5, fill = "steelblue", color = "steelblue") + 
  labs(title = "Violin Plots")
```

If you still want the familiarity of the box plot combined with the enhanced ability to see the underlying distribution you can combine the two plots as well. 

```{r}
p + 
  geom_dotplot(binaxis = "y", stackdir = "center", fill = "steelblue", color = "steelblue") +
  geom_boxplot(alpha = 0, size=1)
```


## <a name="word-clouds"></a> Word Clouds

First we draw a traditional word cloud of the Bertrand Russell's "An essay on the foundations of geometry"

```{r}
library(tm)
library(SnowballC)
library(wordcloud)

setwd("data/")
Russell_Geom <- readChar("Russell_Geometry.txt", file.info("Russell_Geometry.txt")$size)
text_corpus <- Corpus(VectorSource(Russell_Geom)) #Generate a corpus
text_corpus <- tm_map(text_corpus, content_transformer(tolower))
text_corpus <- tm_map(text_corpus, removePunctuation) #remove punctuation
#Remove commonly used words that dont add meaning. (e.g. I, Me)
text_corpus <- tm_map(text_corpus, removeWords, stopwords('english'))
#Turn words into their stems (e.g. Running -> Run)
text_corpus <- tm_map(text_corpus, stemDocument)

wordcloud(text_corpus, max.words = 40, random.order = FALSE)
```

Ahh clearly we can grab very important information on the frequency of the words in this book...

Is "point" or "space" bigger? "Geometry" and "axiom"? Basically it's impossible.

Now let's do it in a bar chart.
```{r}
dtm       <- DocumentTermMatrix(text_corpus)
dtm2      <- as.matrix(dtm)
frequency <- colSums(dtm2)
frequency <- sort(frequency, decreasing=TRUE)

#transform into a tidy dataframe like ggplot desires.
freq_df <- as.data.frame(frequency)
freq_df$word <-  rownames(freq_df)

#sort the data so ggplot respects the dataframe order
ggplot(freq_df[1:40,], aes(x = reorder(word, -frequency), y = frequency)) +
  geom_bar(stat = "identity") +  labs(x = "Word") + #use a barchart and label the xaxis
  theme(axis.text.x = element_text(angle = 40, hjust = 1)) # rotate the text so we can actually read the words
```

So while the bar chart might not be as flashy and cool it certainly more accurately coveys the information that you are trying to show.

That being said, if you are trying to simply make eye candy then go for the word cloud. However, if you are attempting to facilitate meaningful analysis stick to a bar-chart.


---


## <a name="truncated-axes"></a>Trucated Axes

The re-arranging of axis is one of the most potentially damaging forms of data visualization mistakes. By truncating an axis you can entirely change the interpretation of a chart. You can exaggerate a difference or minimize it. A good example of this done recently with potentially dangerous side effects is a tweet sent out by the magazine National Review.

<div style = "text-align: center;">
<img src = "data/climateViz.png" width="500">
</div>

Look at that, we've all been getting way too worried about climate change! But wait, looks like they started their x-axis at 0. Seems like a good idea until you realize that 0 Fahrenheit means absolutely nothing. If you're going to start a temperature at 0 you might as well go all the way and do Kelvin.

Let's see an example of where truncating the axis _is_ bad.

```{r}
data = data.frame("date"  = c(2010, 2011, 2012, 2013),
                  "deaths"= c(400,  425,   430,  440))

plot <- ggplot(data, aes(x = date, y = deaths)) + geom_line() + theme_bw() + labs(title = "Hospital Deaths from 2010-2013")
plot

```

Oh my, looks like we've had a massive spike in hospital deaths.

Deaths however, are a measurement that has a meaningful start point (zero). So let's try and fix out axis scale to represent that.

```{r}
plot + ylim(0,450)
```

Turns out that was a false alarm (although still 40 more deaths might not be trivial).

__Important point:__ Ggplot automatically truncated the axis in this case. In a bar chart it wont let you set a non-zero axis without some esoteric scale commands but for many other plots (such as points and lines) it automatically truncates the axis so your data just fits in the limits. Be vigilant of this.


---


## <a name="multiple-axes"></a>Multiple Axes

<div style = "text-align: center;">
<img src = "data/twoAxes.png" width="700">
</div>

[_source_](http://www.tylervigen.com/spurious-correlations)

Continuing with our morbid theme by looking at Nicholas Cage movies, Tyler Vigen's [excelent site]((http://www.tylervigen.com/spurious-correlations)) on spurious-correlations illustrates our next point very well. When you make a chart with two different axis you can basically make the data say anything you want.

Duke Professor Kieran Healy sums this up very well in a blog post titled ["Two Y-Axes"](https://kieranhealy.org/blog/archives/2016/01/16/two-y-axes/).

<div style = "text-align: center;">
<img src = "data/two_axis.jpg" width="700">
</div>
[_source_](https://kieranhealy.org/blog/archives/2016/01/16/two-y-axes/)

This also goes with the previous point of axes truncation. You can see that by changing axes you can very drastically change interpretations.

Ggplot doesn't even allow multiple axes at all as Hadley Wickham is strongly against the practice.

<div style = "text-align: center;">
<img src = "data/hadley_tweet.png" width="500">
</div>


---


## <a name="information-overload"></a> Information Overload

Say you have a lot of time series data. You might want to compare temporal trends in some measurement for patients in a clinical trial. One natural tendency might be to plot all of their values on the same plot, like below.

```{r}
library(reshape2)
line_data <- data.frame(x_val = 1:50)

for (letter in letters){
  slope = rnorm(1)
  line_data[,letter] <- sin(line_data$x_val + rnorm(1))*slope + rnorm(50)
}

#melt the big dataframe to a tidy one.
tidy_lines = melt(line_data, id = c("x_val"))

#plot with different lines for different letters.
#       data                               color code by letter    draw lines                     add a title
ggplot(tidy_lines, aes(x = x_val, y = value, color = variable)) + geom_line() + labs(title = "Delicious data spagetti")
```

Well this is a mess. You really can't tell what's going on in any way. If you want to see any trends or potential outliers you better be able to distinguish between the shade of green for k and i, and then be able to filter out all the noise and run 50 choose 2 comparisons in your head.

A way to get around this is using a technique known as small multiples. In small multiples you have a bunch of little tiny charts all with a single data element. So in this case it would be 50 separate line plots with one line each.

```{r}
#       data                                   draw lines     seperate charts by the letter           title it
ggplot(tidy_lines, aes(x = x_val, y = value)) + geom_line() + facet_wrap(~variable) + labs(title = "Small multiple lines")
```

As you can see patterns are much easier to see and outliers pop out immediately.

There is another method of dealing with this information overload. Say you have explored your data and want to highlight a single (or maybe two) value in the context of the others. You can highlight that individual line (or whatever graphical element you desire) to call attention to it alone in the chart. This is much more of a explanatory data visualization technique but it does work very well for showing context for an individual element.

```{r}
# cut our dataframe down to just the line we want to show:
z_line <- tidy_lines[tidy_lines$variable == "z", ]

#add the line like the first plot but make them all grey and semi-transparent
ggplot(tidy_lines) + geom_line(aes(x = x_val, y = value, group = variable),color = "grey", alpha = 0.7) +
  labs(title = "Highlighted line") + #Now add a second data element with just highlighted line
  geom_line(data = z_line, aes(x = x_val, y = value, group = variable), color = "steelblue")
```


Sometimes however (Are you sensing a theme?) information overload can be used to your advantage. 

Take for instance the New York Times delegate prediction model: 

<div style = "text-align: center;">
<img src = "data/nyt_delegates.png" width="800">
</div>

[_source_](http://www.nytimes.com/interactive/2016/02/27/upshot/republican-delegate-calculator-how-trump-can-win.html)

In this situation we have a ton of lines, way more than a user can truly parse at one time. In this case this is an intentional method of illustrating uncertainty.  This is a topic that could encompase an entire course. If you are interested in the cutting edge research on uncertianty visualization I suggest you look at Jessica Hulman of the University of Washington's [recent work on the topic.](https://medium.com/hci-design-at-uw/hypothetical-outcomes-plots-experiencing-the-uncertain-b9ea60d7c740#.40rytrmiy) 

_As a side note I feel this is a specific area in which statisticians should be doing the inovating. We understand uncertainty better than most and as is evidenced in Jessica Hulman's post they are representing multiple things the same that perhaps shouldn't be (confidence and credible intervals). If anyone has ideas on this, talk to me!_


---


## <a name="third-dimension"></a> The Third Dimension!

3d charts are cool and very tempting to make, but they are fraught with all sorts of problems. The main one being that perspective (literally) matters. Just like real life, stuff looks bigger the closer it is, so unless your viewer is going to be viewing your visualization on an oculus rift with stereo 3d you should stick to two dimensions. That being said, per usual, there are some ways around this that are acceptable.

With that I give you potentially the worst data visualization ever created:

<div style = "text-align: center;">
<img src = "data/3d_pies_in_the_sky.png" width="500">
</div>

[_source_](http://viz.wtf/)

As we already talked about pie charts are dangerous as slices with different values can look very similar. Once you take that and add in the perspective skewings of the third dimension you get a perfect storm of misleading. I have no suggestions on how to fix this as there are none. Just don't do it. (But later on I will demonstrate an example of when you can use a 3d visualization and be mostly okay.)

---

# <a name="ggplot"></a> Ggplot

While R vs. Python is a heated battle in the statistics community, a much more vitriolic battle is waged on the R sideline over plotting vs base graphics. Jeff Leek's aforementioned article, while written in a tone calling for understanding on the two sides simply ignited passions to hereto unseen levels.  

<div style = "text-align: center;">
<img src = "data/jeffLeekGoogle.png" width="500">
</div>

Ultimately (I get to make this final decision) ggplot has its positives and negatives.

## <a name="strengths"></a>Strengths

- __Has fantastic defaults__

It is pretty dang hard to make a plot with ggplot that looks bad. Base graphics? Pretty easy. This is good as it has helped many people put out better graphics than they otherwise would have.

- __Grammar of Graphics__

the "gg" in ggplot stands for grammar of graphics which is a framework for plotting developed by Leland Wilkinson in his book [The Grammar of Graphics.](http://www.amazon.com/Grammar-Graphics-Statistics-Computing/dp/0387245448) The basic tenants behind this methodology are that you start with your data, and then you assign a geometry to elements of that data, such as circle size to population, then you draw those geometries based upon some scaling of your data. When you think about visualization this way it helps you develop a better understanding of the data itself and think of proper ways to visualize it. (Think the bar vs dot chart.)

- __Verbose__

Due to the grammar of graphics aspect ggplot is rather intelligible. For instance, while writing a line chart takes more characters of code than it does in base graphics it tends to be much clearer what is going on.

```{r, eval=FALSE}
#base graphics
plot(x = df$date, y = df$weight, type = "l", col = "blue")

##ggplot
ggplot(data = df, aes(x = date, y = weight)) + geom_line(color = "blue")
```

Is `col` columns? `type` also seems rather esoteric and would require looking up definitions. In ggplot you can see what x is mapping to, what y is mapping to in your data, `geom_line` is rather clear that it's drawing a line and coloring it blue.

This is important for sharing code with potentially less fluent coders.  

## <a name="weaknesses"></a>Weaknesses

- __Slow__

It generally takes a good bit of time to construct a ggplot graphic. Base allows you to rapidly get a plot up and running. For instance if you want to check if your simulation is running properly or if something interesting is happening in your data a quick `plot(x,y)` is usually more than enough. It doesn't need to look pretty for you.

- __Limitated functionality__

Want to plot a bunch of different charts on a single plot? With ggplot the charts generated with `facet` have to be of the same geometry. If you want to put together a line and bar plot you need to use another library called `grid` which is a pain, especially considering that it's a single simple command it base (`par(mfrow = c(a,b))`).

## <a name="summary"></a>Summary

Like said at the beginning of this document, choose your plotting library and then apply the above principles in it. Very rarely will you need to jump to a whole different library to do something. If you do, that's why stackoverflow was invented. #

---

# <a name="other_libraries"></a>Other Plotting Libraries

Believe it or not there ways to plot data in R other than ggplot and base graphics. The following are a tiny selection of the options. These might not be as fleshed out as ggplot or base but they contain features such as interactivity and larger amounts of automation. 

## <a name="plotly"></a>Plotly

```{r, echo=FALSE}
x <- rnorm(20)
#Function for likelihood
normalLikelihood <- function(x, mu, sigma_2){
  n <- length(x)
  ( (2 * pi * sigma_2)^(-n/2) ) * exp( (-1/(2*sigma_2)) * sum( (x - mu)^2 ))
}

#values to loop over.
mus = seq(mean(x) - 1.5,mean(x) + 1.5, length.out = 500)
sigma_2s = seq(0.001, var(x) + 1.5, length.out = 500)

#how to find where you are on the matrix for grabbing of values.
mu_pos      <- function(i, length){ ( (i-1) %% length ) + 1 }
sigma_2_pos <- function(i, length){ floor((i-1)/length) + 1}

#need how many mus we are looking at.
muLength = length(mus)

likelihoodMat = function(i){
  mu      = mus[mu_pos(i, muLength)]
  sigma_2 = sigma_2s[sigma_2_pos(i, muLength)]
  normalLikelihood(x, mu, sigma_2)
}

#generate the position itterator.
pos = 1: (length(mus)*length(sigma_2s))

results <- sapply(pos, likelihoodMat)

#normalize
newRes = results/max(results)

#make it into a matrix
resMat = matrix(newRes, nrow = muLength)
colnames(resMat) = sigma_2s
rownames(resMat) = mus
```

This is a plotting library that allows you to generate interactive plots directly from R. It does this by rendering them in JavaScript (using a technique we will see shortly).

One beautiful thing about plotly is the ability to export ggplot objects directly to it.

```{r}
library(plotly)

#grab some of R's built in data.
d <- diamonds[sample(nrow(diamonds), 1000), ]

#generate a ggplot object
p <- ggplot(data = d, aes(x = carat, y = price)) +
  geom_point(aes(text = paste("Clarity:", clarity)), size = 2) +
  geom_smooth(aes(colour = cut, fill = cut)) + facet_wrap(~ cut)

#send it to plotly to recreate
ggplotly(p)
```

Now our normal ggplots can have interactivity which can be absolutely fantastic for exploring outliers/ presenting data in an engaging way.

Plotly is not limited simply to re-rendering ggplot. It is capable of rendering three dimensional and/or high performance visualizations using the same engine that video games use.

In the next example `z` is a matrix of data corresponding to a two parameter normal likelihood. We pass it to plotly and tell it to draw a surface plot and ...

```{r, eval=FALSE}
plot_ly(z = resMat, type = "surface")
```

```{r, echo=FALSE}
plot_ly(z = resMat, type = "surface")
```

...pretty cool.

Sometimes a 3d visualization is absolutely necessary (by the data or choice), in these instances you pretty much need the visualization to be interactive (or at least animated) to allow the user to be able to explore the 3d space in order to eliminate the biases injected by perspective.


---

## <a name="html_widgets"></a>HTML-Widgets

An R library that allows you to bind R with JavaScript. Now that much of R is done in rmarkdown notebooks (like this one) and in shiny applications there is a huge potential for interactivity. We have already seen one of these in plotly above but anyone can make them and many have.

Check out the [html widgets showcase](http://www.htmlwidgets.org/showcase_leaflet.html) to see some of the currently made packages. The process for [making one yourself](http://www.htmlwidgets.org/develop_intro.html) is "simple", it's just like making a standard r package except there is also a JavaScript section. JavaScript though is rather... fun.

<a name="manhattan_plot"></a>
I have made a simple htmlwidget library for displaying gwas results and have it hosted on github [here.](https://github.com/nstrayer/D3ManhattanPlots)

```{r}
# library(devtools)
# install_github("nstrayer/D3ManhattanPlots")
library(manhattanPlot) #no idea why I named it different than the repo name

#Load in package data to test.
d = sampleVals

#Draw it!
manhattanPlot(d)  
```

Htmlwidgets are very valuable if you have a particular chart you need to make all the time that would benefit from interactivity and isn't a standard plot like those supported in plotly. More and more researchers are [publishing interactive online visualizations](http://danlarremore.com/faculty/) to go along with their papers and the combination of rmarkdown and htmlwidgets makes this very straightforward. No server needed, simply plop the html page generated by a markdown file anywhere and you're good to go.

__Note:__ If you have something that falls into this category and aren't comfortable with JavaScript visualization get in contact with me. I never turn down an opportunity to make cool visualizations.


---

# <a name="complex_viz"></a>Complex Visualizations

More often than not it is better to keep your visualization simple. However, sometimes it is helpful and or necessary to generate complex visualizations or interactive dashboards. If you have a figure limit for a journal or assignment cramming as much information into a single pane can be very valuable.

## <a name="ggpairs"></a>Ggpairs

There are some methods of doing this with R. Most notable would be making a shiny app with multiple plotting panes. In addition there are some nice packages such as `GGally` which contain functions like `ggpairs` which is a souped up scatter matrix with a bunch of extra information.

```{r}
library(GGally)
data(tips, package = "reshape")
pm <- ggpairs(tips, columns = c(1, 6, 2, 3))

# we have to supress some dumb binwidth warnings that we cant fix because this is a wrapper function.
suppressMessages(print(pm))
```

## <a name="my_stuff"></a>Home Grown Artisan Visualizations

These are some examples of more complex and interactive visualizations I have built for statistics related applications.

- [Confidence Interval Coverage](http://nickstrayer.me/binomialFun/confidenceIntervals)
- [Likelihood Functions](http://nickstrayer.me/likelihood)

## <a name="requirements_analysis"></a>Requirements Analysis

A technique used in software engineering frequently that can be very valuable when designing a complex visualization piece is something know as [requirements analysis.](https://en.wikipedia.org/wiki/Requirements_analysis) Briefly put: the act of sitting down before starting and thoroughly sketching out exactly what you need and want out of the visualization so you don't waste time in the coding process by backtracking.

## <a name="advice"></a>Words To the Wise

Don't go nuts. Most likely the user will not understand the data you are showing at nearly the level you do so something that may seem intuitive to you very well be entirely unintelligible to them.

---

# <a name="beauty_clarity"></a>Beauty Vs. Clarity

A frustratingly qualitative issue in visualization is the role of beauty. Some people such as Edward Tufte are very strict on the matter. Stating that no "chart junk" should clutter your visualization. Basically that only precisely what is necessary to convey the data should be plotted, nothing more. This is very true and most likely the best practice for scientific articles, but many times beauty is a tool for comprehension as well.

If a plot is beautiful the viewer is more likely to stick with it, to investigate it and explore the data within. There is a field of psychology known as [the generation effect](https://en.wikipedia.org/wiki/Generation_effect) which describes how people remember things better that the explore and explain to themselves. If you data is super clearly displayed then the user might simply see it but never remember it.

Take for example the following two visualizations. One is beautiful but certainly doesn't convey the underlying data in the most straightforward way, where as the second is pretty spartan but the data is more than clear. Which one would you spend more time looking at (assuming you're not a physicist)?

<div style = "text-align: center;">
  <img src = "data/prettyViz.jpg" height="450">
  <img src = "data/gravityWaves.jpg" height="450">
</div>
