---
title: "Visualization Practices/Mistakes"
author: "Nick Strayer"
date: "May 17, 2016"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

---

## Outline

- [What to use?](#what-to-use)
- [Common Mistakes/fixes](#common-mistakes)
    - [Pie Charts](#pie-charts)
    - Word Clounds
    - Trucated Axes
    - Multiple Axes
    - Information Overload
    - The Third Dimension
- Ggplot
    - Weaknesses
    - Strengths
    - Bottom Line
- New Tools/Libraries
    - plotly
    - html widgets
      - Manhattan Plots
- Complex Visualizations/Dashboards
    - Ggmatrix
    - Don't go crazy
    - Plan Ahead. 
- Beauty Vs. Clarity. 
  
  ---

## <a name="what-to-use"></a> What to use?

> "The best camera is the one that's with you." –Chase Jarvis.

- Jeff Leek has a [fantastic article](http://simplystatistics.org/2016/02/11/why-i-dont-use-ggplot2/) on his blog about this issue. 

- Ultimately it comes down to what you know. You can do an absolutely amazing amount in most tools (even excel) so do what you like best. 

- For most people in R the choice is Ggplot vs Base. I mostly use Ggplot because it's what I am the most familiar with (and it has nice defaults (more on this later)).

---

## <a name="common-mistakes"></a>Common Mistakes

A lot of data visualization is common sense, but some of it isn't. These are a few of the examples of mistakes made very frequently by visualization practitioners. 

### <a name="pie-charts"></a> The Pie Chart

Okay, let's get the elephant out of the room first. The pie chart elicitaes a similar response in a dataviz person as a computer scientist's prediction algorithm to a statistician. Initially claims of blashpheme but sometimes upon closer inspection grudging respect. 

```{r}
data = data.frame(
  val  = c( 8,   6,   9,   4,   2,  3.5),
  labs = c("a", "b", "c", "d", "e", "f")
)

pie(data$val, data$labs)
```

So why all the ire?

Humans have a very hard time interpreting angles, and that's how a pie chart encodes the data. Looking at the chart above we know that d and f are 0.5 apart, or f is only 87.5% of the size of d, but upon initial inspection the average user would probably say they are the same. 

So let's fix it. 
```{r}
library(ggplot2)
ggplot(data, aes(y = val, x = labs)) + geom_bar(stat = "identity")
```

Now we can clearly see that f is smaller than d

Is it all warrented?

Penn postdoc Randal Olsen has a good [blog post on pie charts.](http://www.randalolson.com/2016/03/24/the-correct-way-to-use-pie-charts/). It is a highly recomended read but 